SI: Settlements

### 2021-04-13 analysis/paper/paper.Rmd
rm dist correlation with kde

---
title: "SI: Settlements"
author:
  - Petr Pajdla:
      email: petr.pajdla@protonmail.com
      institute: [UofO]
      correspondence: false
  - František Trampota:
      email: trampota@phil.muni.cz
      institute: [UofO]
      correspondence: true
institute:
  - UofO: Department of Archaeology and Museology, Masaryk University, Brno
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl"
# abstract: |
keywords: |
  keyword 1; keyword 2; keyword 3
editor_options: 
  markdown: 
    wrap: 80
---

@@

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

```{r packages}
devtools::load_all(".")
library(sf)
library(tidyverse)
```

@@

```{r get_data-rm}
rm_pts <- st_read(here::here(derived_data, "rm_points.geojson"), quiet = TRUE) %>% 
  mutate(label = if_else(orig == "l", "Chipped", "Polished"))
rm_lns <- st_read(here::here(derived_data, "rm_lines.geojson"), quiet = TRUE) %>% 
  mutate(label = if_else(orig == "l", "Chipped", "Polished"))
```

```{r plot-rm, fig.cap = "Distribution of raw material sources."}
ggplot() + 
  geom_sf(data = mask, fill = "gray80", color = NA, alpha = 0.4) +
  geom_sf(data = rivers1, color = "gray60", size = 0.4, alpha = 0.4) +
  geom_sf(data = rivers0, color = "gray60", size = 0.6, alpha = 0.4) +
  geom_sf(data = set_spat, alpha = 0.1, color = "gray80") +
  geom_sf(data = rm_pts, aes(color = rm), shape = 4) +
  geom_sf(data = rm_lns, aes(color = rm)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_wrap(vars(label)) +
  theme_void() +
  theme(panel.border = element_rect(color = "black", fill = NA)) +
  labs(color = "Raw material")
```

@@

rm_dist <- read_csv(paste0(path_temp, "rm_dist.csv"))

rm_dist_median <- rm_dist %>% 
  select(-region, -type, -shp, -min, -mean, -max) %>% 
  group_by(id, rm) %>% 
  summarise(dist_rm = mean(median)) %>% 
  pivot_longer(cols = dist_rm, names_to = "variable") %>% 
  mutate(variable = str_c(variable, rm, sep = "_")) %>% 
  select(-rm)

# update result tibbles
res1_long <- set_base$period1 %>% 
  select(id, period, chrono) %>% 
  inner_join(rm_dist_median) %>%
  bind_rows(res1_long)

res2_long <- set_base$period2 %>% 
  select(id, period, chrono) %>% 
  inner_join(rm_dist_median) %>%
  bind_rows(res2_long)
```

```{r plot-rm-dist-min, fig.cap="Distance to raw material sources."}
rm_dist %>%
  ggplot(aes(x = min, y = rm, fill = region)) +
  geom_violin() +
  scale_fill_brewer(palette = "Accent") +
  facet_wrap(~type, scales = "free_y") +
  labs(x = "min. dist. (km)", y = "raw material", fill = "Region") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r plot-rm-dist-kde-chrono1, fig.cap="KDE against distance to raw material source, Morava river catchment, pottery traditions."}
kde1 %>% select(id, period_label, kde) %>% 
  left_join(rm_dist) %>% 
  mutate(region = str_extract(id, "^.")) %>% 
  filter(rm %in% c("AD", "ZELESICE", "PMD", "KL"),
         region == "M") %>% 
  ggplot(aes(median, kde)) +
  geom_point(alpha = 0.1) + 
  stat_smooth(method = "lm", se = FALSE) +
  facet_grid(vars(period_label), vars(rm)) +
  labs(x = "Median distance to raw material source (km)",
       y = "KDE") +
  theme_minimal() +
  theme(strip.text.y = element_text(angle = 0))
```

```{r plot-rm-dist-kde-chrono2, fig.cap="KDE against distance to raw material source, Morava river catchment, pottery groups."}
kde2 %>% select(id, period_label, kde) %>% 
  left_join(rm_dist) %>% 
  mutate(region = str_extract(id, "^.")) %>% 
  filter(rm %in% c("AD", "ZELESICE", "PMD", "KL"),
         region == "M") %>% 
  ggplot(aes(median, kde)) +
  geom_point(alpha = 0.1) + 
  stat_smooth(method = "lm", se = FALSE) +
  facet_grid(vars(period_label), vars(rm)) +
  labs(x = "Median distance to raw material source (km)",
       y = "KDE") +
  theme_minimal() +
  theme(strip.text.y = element_text(angle = 0))
```

@@

Most of the input data is derived from *DIBAVOD* data set A01 layer
(<https://www.dibavod.cz/>). 
Data for Slovakia are derived from OpenStreet map ? , data for Austria ???

@@

```{r get-raster}
hydro_dens <- raster::raster(
  here::here(derived_data, "hydro", "density_raster.tif")) %>% 
  raster::rasterToPoints() %>%
  as_tibble() %>%
  rename(KDE = density_raster)
```

```{r plot-hydro-density, fig.cap="Kernel smoothed intensity of watercourses overlaid with sites."}
ggplot() +
  geom_raster(data = hydro_dens, aes(x, y, fill = KDE)) +
  geom_sf(data = set_spat, alpha = 0.1, shape = 1, color = "white") +
  theme_void()
```

@@

<!-- The following line inserts a page break -->

>>>>

---
title: "SI: Settlements"
author:
  - Petr Pajdla:
      email: petr.pajdla@protonmail.com
      institute: [UofO]
      correspondence: false
  - František Trampota:
      email: trampota@phil.muni.cz
      institute: [UofO]
      correspondence: true
institute:
  - UofO: Department of Archaeology and Museology, Masaryk University, Brno
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      toc: true
      toc_depth: 3
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl"
# abstract: |
keywords: |
  keyword 1; keyword 2; keyword 3
editor_options: 
  markdown: 
    wrap: 80
---

@@

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

```{r packages, cache=FALSE}
devtools::load_all(".")
library(sf)
library(tidyverse)
library(spatstat)
```

@@

```{r get_data-rm}
rm_pts <- st_read(here::here(derived_data, "rm_points.geojson"), quiet = TRUE) %>% 
  mutate(label = if_else(orig == "l", "Chipped", "Polished"),
         rm = str_to_lower(rm))
rm_lns <- st_read(here::here(derived_data, "rm_lines.geojson"), quiet = TRUE) %>% 
  mutate(label = if_else(orig == "l", "Chipped", "Polished"),
         rm = str_to_lower(rm))
```

```{r plot-rm, fig.cap = "Distribution of raw material sources."}
ggplot() + 
  geom_sf(data = mask, fill = "gray80", color = NA, alpha = 0.4) +
  geom_sf(data = rivers1, color = "gray60", size = 0.4, alpha = 0.4) +
  geom_sf(data = rivers0, color = "gray60", size = 0.6, alpha = 0.4) +
  geom_sf(data = set_spat, alpha = 0.1, color = "gray80") +
  geom_sf(data = rm_pts, aes(color = rm), shape = 4) +
  geom_sf(data = rm_lns, aes(color = rm)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_wrap(vars(label)) +
  theme_void() +
  theme(panel.border = element_rect(color = "black", fill = NA)) +
  labs(color = "Raw material")
```
## Distances

@@

rm_dist <- read_csv(paste0(path_temp, "rm_dist.csv")) %>% 
  mutate(rm = str_to_lower(rm))

# rm_dist_median <- rm_dist %>% 
#   select(-region, -type, -shp, -min, -mean, -max) %>% 
#   group_by(id, rm) %>% 
#   summarise(dist_rm = mean(median)) %>% 
#   pivot_longer(cols = dist_rm, names_to = "variable") %>% 
#   mutate(variable = str_c(variable, rm, sep = "_")) %>% 
#   select(-rm)

rm_dist_mean <- rm_dist %>%
  select(-region, -type, -shp, -min, -median, -max) %>%
  group_by(id, rm) %>%
  summarise(dist_rm = mean(mean)) %>%
  pivot_longer(cols = dist_rm, names_to = "variable") %>%
  mutate(variable = str_c(variable, rm, sep = "_")) %>%
  select(-rm)
```

```{r plot-rm-dist-min, fig.cap="Min. distance to raw material sources."}
rm_dist %>%
  ggplot(aes(x = min, y = rm, fill = region)) +
  geom_violin() +
  scale_fill_brewer(palette = "Accent") +
  facet_wrap(~type, scales = "free_y") +
  labs(x = "min. dist. (km)", y = "raw material", fill = "Region") +
  theme_minimal() +
  theme(legend.position = "bottom")
```
## Settlement density against distance to raw materials

> Hypothesis: In Morava river catchment settlement position is directly associated with 
  raw material source.

We compare settlement density (KDE) with distance to chosen raw materials.
There is no strong evidence that the distribution of settlements is associated with distance to 
any of the key raw material sources occurring in the region in any of the studied periods.
There are some mild negative correlations in the period 4.8 -- 4.2 k BCE for polished stone tools materials.
Strongest connection is observed between given pottery groups in period 4.8 -- 4.6 k BCE in case of ad/pmd/zelesice raw material sources.

```{r plot-rm-dist-kde-chrono1, fig.cap="KDE against distance to raw material source, Morava river catchment, pottery traditions."}
kde1 %>% select(id, period_label, kde) %>% 
  left_join(rm_dist) %>% 
  mutate(region = str_extract(id, "^.")) %>% 
  filter(rm %in% c("ad", "zelesice", "pmd", "kl"),
         region == "M") %>% 
  ggplot(aes(mean, kde)) +
  geom_point(alpha = 0.1) + 
  stat_smooth(method = "lm", se = FALSE) +
  facet_grid(vars(period_label), vars(rm)) +
  labs(x = "Mean distance to raw material source (km)",
       y = "KDE") +
  theme_minimal() +
  theme(strip.text.y = element_text(angle = 0))
```

```{r tab-rm-dist-cor-kde1}
kde1 %>% select(id, period_label, kde) %>% 
  left_join(rm_dist) %>% 
  mutate(region = str_extract(id, "^.")) %>% 
  filter(rm %in% c("ad", "zelesice", "pmd", "kl"),
         region == "M") %>% 
  group_by(period_label, rm) %>% 
  summarise(cor = cor(min, kde)) %>% 
  pivot_wider(id_cols = period_label, names_from = rm, values_from = cor) %>% 
  rename(period = period_label) %>% 
  knitr::kable(caption = "Table of correlations between minimal distance to raw material and settlement density, pottery traditions.") %>% 
  kableExtra::kable_styling(full_width = FALSE)
```

```{r plot-rm-dist-kde-chrono2, fig.cap="KDE against distance to raw material source, Morava river catchment, pottery groups."}
kde2 %>% select(id, period_label, kde) %>% 
  left_join(rm_dist) %>% 
  mutate(region = str_extract(id, "^.")) %>% 
  filter(rm %in% c("ad", "zelesice", "pmd", "kl"),
         region == "M") %>% 
  ggplot(aes(mean, kde)) +
  geom_point(alpha = 0.1) + 
  stat_smooth(method = "lm", se = FALSE) +
  facet_grid(vars(period_label), vars(rm)) +
  labs(x = "Mean distance to raw material source (km)",
       y = "KDE") +
  theme_minimal() +
  theme(strip.text.y = element_text(angle = 0))
```
```{r tab-rm-dist-cor-kde2}
kde2 %>% select(id, period_label, kde) %>% 
  left_join(rm_dist) %>% 
  mutate(region = str_extract(id, "^.")) %>% 
  filter(rm %in% c("ad", "zelesice", "pmd", "kl"),
         region == "M") %>% 
  group_by(period_label, rm) %>% 
  summarise(cor = cor(min, kde)) %>% 
  pivot_wider(id_cols = period_label, names_from = rm, values_from = cor) %>% 
  rename(period = period_label) %>% 
  knitr::kable(caption = "Table of correlations between minimal distance to raw material and settlement density, pottery groups.") %>% 
  kableExtra::kable_styling(full_width = FALSE)
```

## PCA for raw material distances

Principal components analysis is used to reduce dimensionality.
First three PCs are enough to cover more than 95% variability in the data set.

```{r rm-dist-pca}
scale_col <- function(x){
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}

rm_dist_mean_wide <- rm_dist_mean %>%
  mutate(variable = str_remove(variable, "dist_rm_")) %>% 
  pivot_wider(names_from = variable, values_from = value) %>% 
  ungroup() %>% 
  mutate(across(where(is.numeric), scale_col))

# matrix for pca to reduce dimensions
rm_dist_mx <- rm_dist_mean_wide %>% 
  select(-id) %>% 
  as.matrix() 
row.names(rm_dist_mx) <- rm_dist_mean_wide$id

rm_pca <- prcomp(rm_dist_mx)
summary(rm_pca)
```
```{r rm-corrplot, fig.cap="Correlation between principal components and given raw materials."}
corrplot::corrplot(t(rm_pca$rotation[, 1:3]))
```
```{r rm-biplot, fig.cap="First thee principal components (green - East part of Bohemia, blue - Morava river catchment)."}
mycol <- c("B" = "#7FC97F", "M" = "#BEAED4")
pairs(rm_pca$x[, 1:3], col = unname(mycol[str_extract(rownames(rm_dist_mx), ".")]))
```

```{r update-result-rm}
rm_dist_pca <- rm_pca$x[, 1:3] %>% 
  as_tibble(rownames = "id") %>% 
  rename("rm_pc1" = "PC1",
         "rm_pc2" = "PC2",
         "rm_pc3" = "PC3",) %>% 
  pivot_longer(-id, "variable", "value")

# update result tibbles
res1_long <- set_base$period1 %>% 
  select(id, period, chrono) %>% 
  inner_join(rm_dist_pca) %>%
  bind_rows(res1_long)

res2_long <- set_base$period2 %>% 
  select(id, period, chrono) %>% 
  inner_join(rm_dist_pca) %>%
  bind_rows(res2_long)

rm(list = ls(pattern = "^rm_dist_"))
```

@@

Most of the input data is derived from *DIBAVOD* data set A01 layer
(<https://www.dibavod.cz/>).

@@

```{r get-raster}
hydro_dens <- raster::raster(
  here::here(derived_data, "hydro", "density_raster.tif"))

hydro_dens_at_points <- hydro_dens %>% 
  raster::extract(set_spat) %>% 
  as_tibble() %>% 
  rename(hydro_kde = value) %>% 
  bind_cols(select(st_drop_geometry(set_spat), id))

res1_long <- set_base$period1 %>% 
  select(-ends_with("label")) %>% 
  left_join(hydro_dens_at_points) %>% 
  pivot_longer(cols = hydro_kde) %>% 
  rename(variable = name) %>% 
  bind_rows(res1_long)

res2_long <- set_base$period2 %>% 
  select(-ends_with("label"), -facet) %>% 
  left_join(hydro_dens_at_points) %>% 
  pivot_longer(cols = hydro_kde) %>% 
  rename(variable = name) %>% 
  bind_rows(res2_long)
```

```{r plot-hydro-density, fig.cap="Kernel smoothed intensity of watercourses overlaid with sites (raster clipped to research area extent)."}
hydro_dens %>% 
  raster::mask(mask) %>% 
  raster::rasterToPoints() %>%
  as_tibble() %>%
  rename(KDE = density_raster) %>% 
  ggplot() +
  geom_raster(aes(x, y, fill = KDE)) +
  geom_sf(data = set_spat, alpha = 0.1, shape = 1, color = "white") +
  theme_void()
```

# Terrain surrounding the settlements

We focus on slope around the settlements

# Synthesis

@@
